---
output:
  github_document:
    html_preview: true
    toc: false
---


pairscale is a highly optimized and robust implementation for pairwise rescaling that supports various metrics of central tendency, including the mode.

Pairwise normalization consists of two steps. 
First, for every unique pair of columns in the input data matrix, a single measure of central tendency 
(e.g. mean, median, mode) is computed over the difference in their values (i.e. all values in column 
_i_ minus all values in column _j_). The result is a matrix _M_ where columns and rows represent columns
in the input data matrix and values represent their difference (or “distance”). 
Second, we find the optimal scaling factors for each column in the input data matrix such that values in this matrix M are minimized.


&nbsp;

**License:**

[![License: AGPL v3](https://img.shields.io/badge/License-AGPL_v3-blue.svg)](https://www.gnu.org/licenses/agpl-3.0)

**Citation:**

_manuscript under review_



# Installation

## 1) Installing R + tools for compiling

### Windows

- R version 4.1 (or higher) is required on Windows
  - if you already have R installed, check the current version using the R command: `R.version.string`
  - if you need to install or upgrade R
    - download the latest/current version from https://cloud.r-project.org/bin/windows/base/
- RTools is required
  - check RTools is installed using the R command; `pkgbuild::check_rtools()`  (assuming you have the "pkgbuild" package installed)
  - if not, get the RTools installer matching your R version from https://cloud.r-project.org/bin/windows/Rtools/
- Reboot your computer (not mandatory, but this'll avoid obscure bugs/issues)
- Download and install the RStudio desktop IDE. Current link; https://posit.co/download/rstudio-desktop

### MacOS

- R version 4.3 (or higher) is required on MacOS
  - if you already have R installed, check the current version using the R command: `R.version.string`
  - if you need to install or upgrade R
    - download the latest/current version from https://mac.r-project.org . For "apple silicon" MacBooks (approx. 2021 or newer), you'll need "R-`<current version>`-arm64.pkg"
    - don't use the devel/unstable
- Download and install the RStudio desktop IDE. Current link; https://posit.co/download/rstudio-desktop
- Install the toolchain for compiling code
  - start RStudio (if you previously opened it, close and start it to get a 'fresh session')
  - issue the following R statements (documentation @ https://github.com/coatless-mac/macrtools )

```r
# install the macrtools R package
install.packages("remotes")
remotes::install_github("coatless-mac/macrtools")

# non-interactive installation of Xcode CLI
macrtools::xcode_cli_install() 

# install gfortran
macrtools::gfortran_install()

# other binaries required for compiling R
macrtools::recipes_binary_install('r-base-dev')

# install OpenMP and configure Makevars
macrtools::openmp_install()
```

Potential issues;

- `Directory of lock file does not exist: /Users/<username>/Library/Caches/org.R-project.R/R/pkgcache/pkg` during installation suggests you lack write permission to (some part of) this path. Solution; check that this path exists (create if missing) and give your current OS user write permission (at least to the last 4 directories in this path)
- `ld: library not found for -lgfortran` if this is among your error messages, the Fortran/gfortran compiler is not installed
- `Failed to build source package <name>` --> either Xcode/Fortran is not installed, or their version doesn't line up with your current R version

### Linux

- R version 4.1.0 (or higher) and toolchains for compilation are required


## 2) Install the pairscale R package

pairscale is available on CRAN and can be installed using; `install.packages("pairscale")`

- alternatively, install from this GitHub repository instead of CRAN; `remotes::install_github("ftwkoopmans/pairscale")`

On Windows, if you are prompted to install/upgrade RcppArmadillo from source or use a precompiled binary, select "from source".




# Quickstart

- goal: rescale all columns based on pairwise differences such that differences (between columns) are minimized
- input: typically a log2 transformed protein or gene abundance matrix
- this example: a tiny data matrix where we'll use the median-difference between columns as a metric for rescaling/normalizing

```{r}
library(pairscale)
# minimal example; a tiny data matrix
# note how the values in each columns are shifted by a constant
# rescaling should make all columns nearly identical
mat = cbind(
  c(1,2,3,4,5),
  c(2,3,4,5,6),
  c(1.1,2.1,3.1,4.1,5.1),
  c(2,3,4,5,9)
)
print(mat) # input data / prior to normalization

# mat is rescaled and modified in-place by pairscale
# returned variable s represents the applied scaling factors
s = pairscale::pairscale_median(mat, min_value_count = 3)
print(mat) # normalized data
```

Above code rescales matrix `mat` considering all column pairs. But if your dataset has clusters/groups, you can also normalize within cluster/group and then normalize between groups using the `clusters` parameter.

```r
mat_groups = c(1,1,2,2) # provided groups/clusters must be positive integer values
s = pairscale::pairscale_median(mat, min_value_count = 3, clusters = mat_groups)
?pairscale::pairscale_median # see further @ function documentation
```

## a simulated dataset

For many real-world proteomics/transcriptomics datasets, pairwise-differences between samples have an approximately normally distributed null distribution (or more generally, a distribution where the mode is at the center of the real null).
If there are sufficient data points, e.g. at least 50 rows in the input matrix (preferably hundreds), using the mode is a robust metric for determining the relative rescaling of each pair of samples/columns.

- this example: simulate a data matrix with 10 columns/samples, 1000 rows/proteins and a substantial set of missing values for low-abundant rows/proteins in the first 5 samples
- expectation: there are 2 cohorts of distributions, i.e. full distribution and a subset lacking low values, however all overlapping values between columns should be similar post normalization
- observations: 
  - mode normalization is robust and yields expected outcome
  - many data pipelines and algorithms (e.g. directLFQ) normalize between samples using the median value per sample as a reference value ('median shift' approach). This approach is inaccurate for datasets such as generated in this example!


```{r}
library(pairscale)
set.seed(123) # fix random seed for reproducibility
baseline = rnorm(1000, mean = 0, sd = 1) # baseline values for each row/protein
mat = matrix(NA, nrow = 1000, ncol = 10) # simulated data matrix
simulated_scaling_factors = runif(n = 10, min = -3, max = 3) # randomly rescale each column
for(j in 1:10) {
  # values in column j; baseline values + a bit of random noise per row + mock scaling factor
  mat[,j] = baseline + runif(n = 1000, min = -0.1, max = 0.1) + simulated_scaling_factors[j]
  # in the first 5 samples, remove the lowest 25% of values (e.g. strong censoring)
  # as a result, the distributions between the first and second set of samples are distinct
  if(j <= 5) {
    mat[head(order(mat[,j], decreasing = FALSE), n = 250),j] = NA
  }
}

# QC: plot the data distributions
plot(NA, xlim = c(-4, 4), ylim = c(0, 0.75), type = "n", xlab = "value", ylab = "density", main = "INPUT DATA\nblack = full distribution in columns 6:10\nred = subset in columns 1:5")
for(j in 1:ncol(mat)) {
  lines(density(mat[,j], na.rm = TRUE), col = 1 + (j <= 5))
}


# use pairscale to apply mode normalization
# variable 'mat' is updated by reference (no overhead of creating a copy of the data matrix)
# returned variable represents the applied scaling factors
estimated_scaling_factors = pairscale::pairscale_mode(mat, min_value_count = 25)

# for reference, compute scaling factors using the commonly used 'median shift' approach
mat_column_medians = apply(mat, 2, median, na.rm = TRUE)

# compare simulated differences in 'sample loading' against estimated normalization factors
# 1) pairscale recapitulates expected scaling factors; minor imprecision due to simulated random noise
# 2) column medians are not proper scaling factors for this simulated dataset;
#     'median shift' works for homogenous datasets but fails when sample distributions are dissimilar
print(cbind(
 "simulated" = simulated_scaling_factors - mean(simulated_scaling_factors), 
 "pairscale estimated" = estimated_scaling_factors - mean(estimated_scaling_factors),
 "column medians" = mat_column_medians - mean(mat_column_medians) 
))


plot(NA, xlim = c(-4, 4), ylim = c(0, 0.75), type = "n", xlab = "value", ylab = "density", main = "NORMALIZED\nblack = full distribution in columns 6:10\nred = subset in columns 1:5")
for(j in 1:ncol(mat)) {
  lines(density(mat[,j], na.rm = TRUE), col = 1 + (j <= 5))
}

# visualize the normalized data matrix to illustrate all rows are similar except those set to NA
heatmap(mat[order(baseline),], Rowv = NA, Colv = NA, scale = "none")
```




# benchmark computation speed

Here we evaluate the computation speed of pairscale functions that compute measures of central tendency for a given numeric vector;

```{r eval=FALSE}
library(pairscale)
if(!requireNamespace("rbenchmark", quietly = TRUE)) install.packages("rbenchmark")

# find the mode using base R density function
r_density_mode = function(x, trim = 0) {
  if(trim > 0) {
    q = quantile(x, probs = c(trim, 1-trim), na.rm = TRUE)
    x = x[is.finite(x) & x > q[1] & x < q[2]]
  }
  d = stats::density(x, adjust = 1, n = 512, bw = "nrd", na.rm = TRUE)
  return(d$x[which.max(d$y)])
}

# 100k vector + 1 extreme value
x = c(rnorm(100000, mean = 0, sd = 1), 13)

# use rbenchmark package to time pairscale functions for computing measures of central tendency
# benchmark results shown here were generated on a MacBook Pro with a Apple M4 Max chip under R 4.4.2
rbenchmark::benchmark(
  "R trimmedmean" = {d = mean(x, trim = 0.2, na.rm=T)},
  "R mode" = {d = r_density_mode(x)},
  "R mode trimmed" = {d = r_density_mode(x, trim = 0.025)},
  "R median" = {d = median(x, na.rm=T)},
  "vector_median" = {d = vector_median(x)},
  "vector_trimmedmean" = {d = vector_trimmedmean(x, trim = 0.2)},
  "vector_mode(fastest 200)" = {d = vector_mode(x, min_value_count = 1, adjust = 1, kernel_width_in_sd = 3, n_bins = 200, bandwidth_method = "nrd_fastest")},
  "vector_mode(fast 200)" = {d = vector_mode(x, min_value_count = 1, adjust = 1, kernel_width_in_sd = 3, n_bins = 200, bandwidth_method = "nrd_fast")},
  "vector_mode(200)" = {d = vector_mode(x, min_value_count = 1, adjust = 1, kernel_width_in_sd = 3, n_bins = 200, bandwidth_method = "nrd")},
  "vector_mode(fastest 512)" = {d = vector_mode(x, min_value_count = 1, adjust = 1, kernel_width_in_sd = 3, n_bins = 512, bandwidth_method = "nrd_fastest")},
  "vector_mode(fast 512)" = {d = vector_mode(x, min_value_count = 1, adjust = 1, kernel_width_in_sd = 3, n_bins = 512, bandwidth_method = "nrd_fast")},
  "vector_mode(512)" = {d = vector_mode(x, min_value_count = 1, adjust = 1, kernel_width_in_sd = 3, n_bins = 512, bandwidth_method = "nrd")},
  replications = 1000, order = "relative", columns = c("test", "replications", "elapsed", "relative", "user.self", "sys.self")
)
```

Results on a MacBook Pro:

```
#                        test replications elapsed relative user.self sys.self
# 7  vector_mode(fastest 200)         1000   0.195    1.000     0.194    0.000
# 10 vector_mode(fastest 512)         1000   0.203    1.041     0.202    0.000
# 5             vector_median         1000   0.474    2.431     0.471    0.003
# 6        vector_trimmedmean         1000   0.782    4.010     0.782    0.000
# 4                  R median         1000   0.984    5.046     0.828    0.155
# 11    vector_mode(fast 512)         1000   0.999    5.123     0.998    0.001
# 8     vector_mode(fast 200)         1000   1.059    5.431     1.058    0.001
# 12         vector_mode(512)         1000   1.347    6.908     1.345    0.002
# 9          vector_mode(200)         1000   1.453    7.451     1.451    0.002
# 1             R trimmedmean         1000   1.479    7.585     1.265    0.213
# 2                    R mode         1000   2.012   10.318     1.831    0.180
# 3            R mode trimmed         1000   3.749   19.226     3.294    0.454
```

Results on a Windows workstation:

```
#                        test replications elapsed relative user.self sys.self
# 10 vector_mode(fastest 512)         1000    0.38    1.000      0.38     0.00
# 7  vector_mode(fastest 200)         1000    0.43    1.132      0.43     0.00
# 5             vector_median         1000    0.72    1.895      0.72     0.00
# 6        vector_trimmedmean         1000    1.14    3.000      1.14     0.00
# 11    vector_mode(fast 512)         1000    1.30    3.421      1.30     0.00
# 8     vector_mode(fast 200)         1000    1.33    3.500      1.33     0.00
# 12         vector_mode(512)         1000    1.75    4.605      1.75     0.00
# 9          vector_mode(200)         1000    1.76    4.632      1.76     0.00
# 4                  R median         1000    2.09    5.500      2.01     0.08
# 1             R trimmedmean         1000    4.24   11.158      3.14     1.11
# 2                    R mode         1000    4.86   12.789      4.06     0.80
# 3            R mode trimmed         1000    7.71   20.289      7.39     0.33
```


## a note for R package developers

If you download and open the pairscale.Rproj file and use `devtools::load_all()`,
instead of going the usual route of `install.packages("pairscale"); library(pairscale)`,
the above benchmarks will underperform because devtools will override compiler flags (setting e.g. -O0).
To disable this, and thus only use your ~/.R/Makevars (if available), run this code instead;
```r
Sys.setenv("PKG_BUILD_EXTRA_FLAGS" = "false")
devtools::load_all(recompile = TRUE)
```
